---
title: "UPDATED Overton Cleaning (Markdown)"
author: "Khristianne Ermino"
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: kable
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts = list(width.cutoff = 60), tidy=TRUE)
```

```{r}
library(tidyverse)
```

# Loading In Reports CSV
```{r}
df <-read_delim("overton_withSara/reports.csv")
```

# Creating a Test Trial
```{r}
test <-df[1:5,] %>% 
  select ("Title of citing document") %>%
  rename(title = "Title of citing document")
```

Rather than figuring out code for all ~26,000 entries, it would be easier to work on it with a smaller test set. Therefore, the first 5 entries from the database were chosen from the "Title of citing document" column. This long titles was renamed as "title" for the sake of efficiency when coding. 

# Testing Regex for String Extraction and Create a New Column with the Date
```{r}
str_extract(test$title,"(\\d\\d\\d\\d)+(?=E)")
```

Using the str_extract function, we are accessing the column "title" from the data frame called "test". This allows us to extract the 4 wildcard digits that occur before the "E". Since we're working with a test set of the first 5 entries, we know that they should all be "2010". This allows us to check if the code has worked in the way we want it to. 

```{r}
test %>% 
  mutate(year = str_extract(test$title,"(\\d\\d\\d\\d)+(?=E)")) 
```

Using the mutate function alongside the str_extract function, the 4 wildcard digits that are before the "E" from the title in the "test" data frame is taken and mutated to create a new column in the "test" data frame called "year". 

# Testing the Separate Function for Species Name Pull
```{r}
test %>%
  separate(title, into = c(NA, "int1"), sep = "COSEWIC assessment and status report on the ") %>% 
  separate(int1, into = c("full.spp", NA), sep = ", in Canada") %>%
  separate(full.spp, into = c("Common", "Latin"), sep = ', ')
```

In the title column of the "test" data frame, the contents are separated using the separate function. The separator is "COSEWIC..." with the contents prior to the separator being ignored  while the contents after the separator are placed into a column labelled "int1" (NA, "int1"). 

The separate function is used once again, with the contents of the int1 column being separated using ", in Canada" as the separator. The contents prior to the separator are placed into a column labelled "full.spp" while the contents prior are ignored. 

Using the separate function, the contents of the "full.spp" column is separated with the separator being ", ", keeping the contents of both sides of the separator with the contents prior placed in the "Common" column and the contents after the separator placed in the "Latin" column.

# Full df Clean 
```{r}
df.pub.year <- df %>% 
  rename(title = "Title of citing document", pub.date = "Cited published on date") %>%
  separate(pub.date, into = c("pub.year"), sep = '-', extra = "drop") 
```

Cleaning title names and separating the year of publication from the "pub.date" column. From the loaded overton database called "df', the long titles were renamed for more efficient coding. Separates the first part of the "pub.date" using the "-" separator and drops the other parts. 

## Making Intermediate Obj So That the String Extract Works Properly

```{r}
int.year <- df.pub.year %>%
  mutate(year = str_extract(df.pub.year$title,"(\\d\\d\\d\\d)+(?=E)")) %>%
  separate(`Published on date of citing document`, into = c("cit_doc_year"),
    sep = "-", extra = "drop") %>%
  mutate(year = if_else(is.na(year), cit_doc_year, year))
```

Using the str_extract, from the "title" column of the df.pub.year" data frame, take out the four wildcard digits that are found before the "E". Using the mutate function, a "year" column is created in the "int.year" data frame.

