---
title: "UPDATED Cleaning DB Files (Markdown)"
author: "Khristianne Ermino"
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: kable
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts = list(width.cutoff = 60), tidy=TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

# Loading cleanDB.csv File
```{r}
df <- read_delim("finalcleanDB_(03-05-26).csv") %>%
  mutate(common = str_to_title(common))
View(df)
```

The "common" species name was converted to title case for consistency.

# Grouping Common and Latin Names with NAs
```{r}
na_commonandlatin <- df %>%
  filter(is.na(common), is.na(latin)) %>% #Filtered for species that contain NAs in their common and latin columns. 
  group_by(pub.title) %>% #Grouped them by the pub.title column. 
  summarise(count = n())
head(na_commonandlatin) 
```

No entries! No entries were missed in the manual placement of data. 
Double Checking. 
```{r}
figureplot1 <- df %>%
  mutate(source_type = ifelse(is.na(source), "Grey Literature", "Academic Journal"))
View(figureplot1)

year_count1 <-figureplot1 %>%
  group_by(final.year, source_type) %>%
  summarise(n = n(), .groups = "drop")
View(year_count1)

ggplot(year_count1, aes(x = final.year, y = n, fill = source_type)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Year of publication",
    y = "Number of citations",
    fill = "Information source") +
  scale_fill_manual(values = c(
    "Academic Journal" = "#0072B2",
    "Grey Literature" = "#E69F00")) + theme_classic() 
```

```{r}
df %>%
  filter(common == "American Eel") %>%
  distinct(final.year)
```

# Selecting the Most Recent Assessments
```{r}
#Wanted to check if there's a more efficient way. Found a function online that selects the max value. Used on "year" column while grouping "common" column. 
cleanestdb <- df %>%
  group_by(common) %>%
  slice_max(order_by = final.year) #Found function online that allows us to select max value. 
View(cleanestdb) 
```

```{r}
df %>% 
  filter(is.na(journal.pub.year)) %>%
  count() #129

missing_year_df <- df %>%
  filter(is.na(journal.pub.year))
View(missing_year_df)
```

#Shows how many that 129 entries don't have a value in the journal.pub.year
Should we remove them? - YES.

Removing NAs from journal.pub.year
```{r}
df_cleanforfigure <- df %>%
  filter(!is.na(journal.pub.year))
View(df_cleanforfigure)

df_cleanforfigure %>%
  filter(is.na(journal.pub.year)) %>%
  count()
View(df_cleanforfigure)
```

```{r}
write_delim(df_cleanforfigure, "overton_withSara/df_cleanforfigure.csv", delim = ',')
```

Creating the Stacked Bar Plot - AFTER MOST RECENT
```{r}
figureplot <- df_cleanforfigure %>%
  mutate(source_type = ifelse(is.na(source), "Grey Literature", "Academic Journal"))
View(figureplot)

year_count <-figureplot %>%
  group_by(final.year, source_type) %>%
  summarise(n = n(), .groups = "drop")
View(year_count)

ggplot(year_count, aes(x = final.year, y = n, fill = source_type)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Year of publication",
    y = "Number of citations",
    fill = "Information source"
  ) +
  scale_fill_manual(values = c(
    "Academic Journal" = "#0072B2",
    "Grey Literature" = "#E69F00"
  )) +
  theme_classic() 
```

Need to remove NA source for next steps. 
```{r}
df_clean <- cleanestdb %>%
  filter(!is.na(source))
```
Removed 680 entries. Are all of those entries grey literature? 

```{r}
mostcleanestdb <- na.omit(cleanestdb)
View(mostcleanestdb)
# Omitted all rows with NAs (Double checked everything within column and latin columns)